Efficient Handling of String-Number Conversion

### 2019-11-22 draft.tex
abstract
related work and minor update to evalutaion

\begin{abstract}
String-number conversion is an important class of constraint needed for the symbolic execution for programs. It is particular important in the analysis of scripting languages such as JavaScript and Python, where string-number conversion is a part of the definition of the core semantics of the language. However, the state-of-the-art constraint solvers have only very limited support to this type of constraint. In this paper, we propose an approach that can support a combination of string-number conversion as well as other common types of string constraints efficiently. Experimental results show that it significantly outperforms other state-of-the-art tools on benchmarks that involves string-number conversion.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\hide}[1]{}
\newcommand{\tool}{{\textsf{Z3-PFA}}}
\newcommand{\dr}{\mathcal{R}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\changed}[1]{{\color{orange}#1}}
\newcommand{\yfc}[1]{{\color{blue}YFC: #1}}
\newcommand{\petr}[1]{{\color{pink}Petr: #1}}
\newcommand{\chatAt}[2]{\mbox{\textsf{charAt}($#1$, $#2$)}}
\newcommand{\ite}[3]{\mbox{\textsf{ite}($#1$, $#2$, $#3$)}}
\newcommand{\sti}[1]{\mbox{\textsf{toNum}($#1$)}}
\newcommand{\its}[1]{\mbox{\textsf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvars}{\mbox{$\mathbb{V}_{\Sigma_\epsilon}$}}
\newcommand{\pvars}{\mbox{$\mathbb{V}_{\sharp}$}}
%\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvarone}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvartwo}{\mbox{$V'_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{V}
\newcommand{\cvartwo}{V'}
\newcommand{\cvar}{V}
\newcommand{\modelsof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
%\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}
\newcommand{\parikhof}[1]{\mathbb{P}{(#1)}}
\newcommand{\parikhwof}[2]{|#1|_{#2}}
%\newcommand{\semof}[1]{||#1||}
\newcommand{\semof}[1]{\modelsof{#1}}
\newcommand{\decode}[1]{\mathit{decode_{#1}}}
%\newcommand{\parikhfof}[1]{\Phi_{\parikhof{#1}}}
\newcommand{\parikhfof}[1]{\Phi_{\mathbb{P}}(#1)} %shouldn't it be something that takes automaton as parameter and returns formula?
\newcommand{\pim}{I_{\#}}
%\newcommand{\syncop}{\Cap}
\newcommand{\syncop}{\times}
\newcommand{\syncof}[2]{#1 \syncop #2}
\newcommand{\syncfof}[2]{\Psi_{\syncof {#1} {#2}}}
\newcommand{\syncT}{T_\syncop}
\newcommand{\pvarsof}[1]{\#{#1}}
\newcommand{\pvar}{\pvarsof V}
\newcommand{\pvarone}{\pvarsof V}
\newcommand{\pvartwo}{\pvarsof {V'}}
\newcommand{\defeq}{::=}
\newcommand{\encode}[1]{\mathit{encode}_{#1}}
\newcommand{\eqwrt}[1]{=_{#1}}
\newcommand{\iequiv}{\equiv}
\newcommand{\restrict}[2]{#1_{#2}}
\newcommand{\underf}[2]{\mathit{under}_#1(#2)}
\newcommand{\paf}{\psi}

@@

%As solvers are applied to symbolic execution of programming languages, this is increasingly becoming a severe limitation. For many languages such as Java, conversion between integers and strings is a niche operation, and much symbolic execution can be done without supporting it; however, this is not true of some scripting languages, of which JavaScript is the most notable. JavaScript is a key language---it powers most interactive content on the Web, and increasingly server-side code with Node.js---and it has conversions between strings and integers embedded in the core of its semantics. 
\smallskip
\begin{center}
	\begin{minipage}{5cm}
		\begin{verbatim}
    for(var i = 0; i < 10; i++) {
        arr[i] = 0;
   }
		\end{verbatim}
	\end{minipage}
\end{center}
\smallskip

@@

Assuming that the domain restriction for $x$ is $(A^m,\paf^m)$, 
the value of the integer $n$ can be extracted from a numeral using the formula
$$
\Psi_{\mathsf{toInt}}^m\defeq \bigvee_{1\leq k \leq m} \Psi_{\mathsf{last(k)}}^m \land (n = v_1* 10^{k-1} + v_2* 10^{k-2} + \ldots + v_{k})
$$
where $\Psi_{\mathsf{last(k)}}^m$ says that the last variable of $A^m$ assigned a non-$\epsilon$ value is $v_{k}$, namely 
$$
\Psi_{\mathsf{last(k)}}^m \defeq (k = m \land v_k \geq 0) \lor (v_{k} \geq 0 \land v_{k+1} = -1)\ .
$$
Since we also need to distinguish the case when $x$ is not a number, in which case $n$ should equal $-1$, 
the formula underapproximating $\phi_s$ is finally constructed as  
\begin{multline*}
\underf \dr {\phi_s} 
\defeq 
\parikhfof{A^m} 
\land \\
\bigl((\Psi_{\mathsf{NaN}}^m \land n = -1) 
\lor 
(\neg\Psi_{\mathsf{NaN}}^m \land \Psi_{\mathsf{toInt}}^m)\bigr)
\end{multline*}
The following lemma states correctness of this construction:
\begin{lemma}
$
\semof {\underf \dr {\psi_s} }_{V_\dr \cup \#V_\dr \cup \{n\}} = \encode  \dr(\semof{\phi_s})
$
\end{lemma}

@@

\paragraph{Related works:} 
Already in 1946, Quine \cite{quine1946concatenation} showed that the first order theory
of string equations is undecidable.
%
An important line of work has been to identify subclasses
for which decidability can be achieved.
%
The pioneering work by Makanin \cite{makanin1977problem} proposed a decision
procedure for quantifier-free word equations, i.e., Boolean combinations of
equalities and disequalities, where the variables may denote words of 
arbitrary lengths.
%
The decidability and complexity of different subclasses
have been considered by several works, e.g.
\cite{plandowski1999satisfiability,plandowski2006efficient,matiyasevich2008computation,robson1999quadratic,schulz1990makanin,ganesh2012word,ganesh2016undecidability}.
Generalizations of the work of Makanin by adding
new types of constraints have been difficult to achieve.
%
For instance, the satisfiability of word equations combined with length
constraints of the form $\left|x\right|=\left|y\right|$ is open
\cite{buchi1990definability}.
%
Recently, 
regular and especially relational transducers constraints were identified
%
as a strongly desirable feature of string languages 
%
especially in the context software analysis with an emphasis on security. 
%
Adding these to the mix leads immediately to undecidability \cite{morvan2000rational}
%
and hence numerous decidable fragments were proposed \cite{abdulla2014string,barcelo2013graph,lin2016string,chen2017decidable,chen2019decision}. 
%
From these, the straight line fragment of \cite{lin2016string} is the most general decidable combination of concatenation and transducers. It is however incomparable to the acyclic fragment of \cite{abdulla2014string} (which does not have transducers but could be extended with them in a straightforward manner).
%
Some works add also other syntactic features, such as \cite{chen2017decidable,chen2019decision}, 
but the limit of decidable combinations of the core string features---transducers/regular constraints, length constraints, and concatenation stays at \cite{lin2016string} and \cite{abdulla2014string}.
%
The weakly chaining decidable fragment present in this paper significantly generalises both these fragments in a practically relevant direction.

The strong practical motivation in string solving led to a rise of a number of SMT solvers that do not always provide completeness guarantees but concentrate on solving practical problem instances, 
through applying a variety of calculi and algorithms.
%
A number of tools handle string constraints by means of {\it length-based under-approximation} and
translation to
bit-vectors~\cite{kiezun2009hampi,saxena2010symbolic,saxena2010flax}, assuming
a fixed upper bound on the length of the possible solutions. 
%
Our method on the other hand allows to analyse constraints without a length limit and with completeness guarantees.
%
More recently, also {\it DPLL(T)-based} string solvers lift the
restriction of strings of bounded length; this generation of solvers
includes Z3-str3 \cite{berzish2017z3str3}, Z3-str2~\cite{zheng2017z3str2}, 
CVC4~\cite{cvc4Tool}, S3P~\cite{trinh2014s3,trinh2016progressive},
Norn~\cite{abdulla2015norn}, 
Trau~\cite{abdulla2018trau}, 
Sloth~\cite{sloth}, 
and 
Ostrich~\cite{chen2019decision}. DPLL(T)-based solvers
handle a variety of string constraints, including word equations, regular expression membership, length constraints, and (more rarely)
regular/rational relations; the solvers are not complete for the full
combination of those constraints though, and often only decide a (more
or less well-defined) fragment of the individual constraints. 
Equality constraints are normally handled by means of splitting into simpler
sub-cases, in combination with powerful techniques for Boolean
reasoning to curb the resulting exponential search space. 
%
Our implementation is combining strong completeness guarantees of \cite{sloth} extended to handle the fragment proposed in this paper with an efficient approximation techniques of \cite{abdulla2018trau} 
and its performance on existing benchmarks compares favourably with the most efficient of the above tools.
%

A further direction is {\it automata-based} solvers for analyzing
string-manipulated programs. Stranger~\cite{yu2010stranger} soundly
over-approximates string constraints using regular languages, and
outperforms DPLL(T)-based solvers when checking single execution
traces, according to some evaluations~\cite{kausler2014evaluation}. It has
recently also been observed~\cite{wang2016string,sloth} that
automata-based algorithms can be combined with model checking
algorithms, in particular IC3/PDR, for more efficient checking of the emptiness for automata. However, many kinds of constraints, including length
constraints and word equations, cannot be
handled by automata-based solvers in a complete manner.

>>>>

\begin{abstract}
String-number conversion is an important class of constraint needed for the symbolic execution for programs. It is particular important in the analysis of scripting languages such as JavaScript and Python, where string-number conversion is a part of the definition of the core semantics of the language. However, the state-of-the-art constraint solvers have only very limited support to this type of constraint. In this paper, we propose an approach that can support a combination of string-number conversion as well as other common types of string constraints efficiently. Experimental results show that it significantly outperforms other state-of-the-art tools on benchmarks that involves string-number conversion.
\end{abstract}
 \setcopyright{none} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\hide}[1]{}
\newcommand{\tool}{{\textsf{Z3-PFA}}}
\newcommand{\dr}{\mathcal{R}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\changed}[1]{{\color{orange}#1}}
\newcommand{\yfc}[1]{{\color{blue}YFC: #1}}
\newcommand{\petr}[1]{{\color{pink}Petr: #1}}
\newcommand{\chatAt}[2]{\mbox{\textsf{charAt}($#1$, $#2$)}}
\newcommand{\ite}[3]{\mbox{\textsf{ite}($#1$, $#2$, $#3$)}}
\newcommand{\sti}[1]{\mbox{\textsf{toNum}($#1$)}}
\newcommand{\its}[1]{\mbox{\textsf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvars}{\mbox{$\mathbb{V}_{\Sigma_\epsilon}$}}
\newcommand{\pvars}{\mbox{$\mathbb{V}_{\sharp}$}}
%\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvarone}{\mbox{$V_{\Sigma_\epsilon}$}}
%\newcommand{\cvartwo}{\mbox{$V'_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{V}
\newcommand{\cvartwo}{V'}
\newcommand{\cvar}{V}
\newcommand{\modelsof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
%\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}
\newcommand{\parikhof}[1]{\mathbb{P}{(#1)}}
\newcommand{\parikhwof}[2]{|#1|_{#2}}
%\newcommand{\semof}[1]{||#1||}
\newcommand{\semof}[1]{\modelsof{#1}}
\newcommand{\decode}[1]{\mathit{decode_{#1}}}
%\newcommand{\parikhfof}[1]{\Phi_{\parikhof{#1}}}
\newcommand{\parikhfof}[1]{\Phi_{\mathbb{P}}(#1)} %shouldn't it be something that takes automaton as parameter and returns formula?
\newcommand{\pim}{I_{\#}}
%\newcommand{\syncop}{\Cap}
\newcommand{\syncop}{\times}
\newcommand{\syncof}[2]{#1 \syncop #2}
\newcommand{\syncfof}[2]{\Psi_{\syncof {#1} {#2}}}
\newcommand{\syncT}{T_\syncop}
\newcommand{\pvarsof}[1]{\#{#1}}
\newcommand{\pvar}{\pvarsof V}
\newcommand{\pvarone}{\pvarsof V}
\newcommand{\pvartwo}{\pvarsof {V'}}
\newcommand{\defeq}{::=}
\newcommand{\encode}[1]{\mathit{encode}_{#1}}
\newcommand{\eqwrt}[1]{=_{#1}}
\newcommand{\iequiv}{\equiv}
\newcommand{\restrict}[2]{#1_{#2}}
\newcommand{\underf}[2]{\mathit{under}_#1(#2)}
\newcommand{\paf}{\psi}

@@

%As solvers are applied to symbolic execution of programming languages, this is increasingly becoming a severe limitation. For many languages such as Java, conversion between integers and strings is a niche operation, and much symbolic execution can be done without supporting it; however, this is not true of some scripting languages, of which JavaScript is the most notable. JavaScript is a key language---it powers most interactive content on the Web, and increasingly server-side code with Node.js---and it has conversions between strings and integers embedded in the core of its semantics.

\begin{center}
	\begin{minipage}{5cm}
		\begin{verbatim}
    for(var i = 0; i < 10; i++) {
        arr[i] = 0;
   }
		\end{verbatim}
	\end{minipage}
\end{center}

@@

Assuming that the domain restriction for $x$ is $(A^m,\paf^m)$, 
the value of the integer $n$ can be extracted from a numeral using the formula
$$
\Psi_{\mathsf{toInt}}^m\defeq \bigvee_{1\leq k \leq m} \Psi_{\mathsf{last(k)}}^m \land (n = v_1* 10^{k-1} + v_2* 10^{k-2} + \ldots + v_{k})
$$
where $\Psi_{\mathsf{last(k)}}^m$ says that the last variable of $A^m$ assigned a non-$\epsilon$ value is $v_{k}$, namely 
$$
\Psi_{\mathsf{last(k)}}^m \defeq (k = m \land v_k \geq 0) \lor (v_{k} \geq 0 \land v_{k+1} = -1)\ .
$$
Since we also need to distinguish the case when $x$ is not a number, in which case $n$ should equal $-1$, 
the formula under-approximating $\phi_s$ is finally constructed as  
\begin{multline*}
\underf \dr {\phi_s} 
\defeq 
\parikhfof{A^m} 
\land \\
\bigl((\Psi_{\mathsf{NaN}}^m \land n = -1) 
\lor 
(\neg\Psi_{\mathsf{NaN}}^m \land \Psi_{\mathsf{toInt}}^m)\bigr)
\end{multline*}
The following lemma states correctness of this construction:
\begin{lemma}
$
\semof {\underf \dr {\psi_s} }_{V_\dr \cup \#V_\dr \cup \{n\}} = \encode  \dr(\semof{\phi_s})
$
\end{lemma}

@@

\section{Related works} 
To the best of our knowledge, the study of solving string constraint traces back to 1946, when Quine \cite{quine1946concatenation} showed that the first-order theory of string equality constraints (a.k.a. word equations) is undecidable. Makanin achieves a milestone \cite{makanin1977problem} by showing that the class of quantifier-free string equality constraints is decidable.
Since then, several works, e.g., \cite{plandowski1999satisfiability,plandowski2006efficient,matiyasevich2008computation,robson1999quadratic,schulz1990makanin,ganesh2012word,ganesh2016undecidability}, consider the decidability and complexity of different subclasses of string equality constraints.

Satisfiability of string constraints is a challenging problem. The satisfiability of equality constraints combined with length constraints of the form $|x|=|y|$ is already opened for more than 20 years \cite{buchi1990definability}.
Numerous decidable fragments were proposed\cite{abdulla2014string,barcelo2013graph,lin2016string,chen2017decidable,chen2019decision,abdulla2019chain}. Among them, the chain-free fragment \cite{abdulla2019chain} used by our over-approximation module is the largest known decidable fragment, which allows us to produce more precise over-approximation and hence solve many UNSAT instances efficiently.

The strong practical motivation led to the rise of several string constraints solvers that concentrate on solving practical problem instances. Several tools handle string constraints assuming a fixed upper bound on the length of strings and translate them to boolean satisfiability problems~\cite{kiezun2009hampi,saxena2010symbolic,saxena2010flax}.  Our method, on the other hand, allows analyzing constraints without a length limit and still with some completeness guarantees, i.e., within the language defined by PFAs.

More recently,  {\it DPLL(T)-based} string solvers~\cite{berzish2017z3str3,zheng2017z3str2,cvc4Tool,trinh2014s3,trinh2016progressive,abdulla2015norn,abdulla2018trau,sloth,chen2019decision} lift the
restriction of strings of bounded length. They usually support a variety of string constraints, including all basic string constraints, and sometimes also regular/rational relations. 
The typical procedure they used for solving equality constraints is to split them into simpler sub-cases, in combination with powerful techniques for Boolean reasoning to curb the resulting exponential search space. In contrast, our approach uses a completely different search strategy. We restrict the solution space to some predefined pattern and step-wisely enrich the pattern in use.

The most related work to ours is~\cite{abdulla2018trau} who proposed to project the solution space of variables to a generalization of flat automata. The main difference is that our approach works fully symbolically, thanks to the ideas of using character variables. The use of character variables enables our approach to handle string-number constraints efficiently -- the values and number of occurrences of those variables can be directly convert to a number in a linear formula. Moreover, the approach of~\cite{abdulla2018trau} is exponential to the alphabet size, and hence in the implementation, usually requires heuristics to prune out unnecessary symbols.

A further direction is {\it automata-based} solvers for analyzing
string-manipulated programs. ABC~\cite{aydin2018parameterized} and Stranger~\cite{yu2010stranger} soundly over-approximates string constraints using multi-tape automata~\cite{yu2016optimal}, and outperforms DPLL(T)-based solvers when checking single execution traces, according to some evaluations~\cite{kausler2014evaluation}. People also studied the combination of automata-based algorithms with with model checking
algorithms, in particular, IC3/PDR, for more efficient checking of the emptiness for automata~\cite{wang2016string,sloth}. However, many kinds of constraints, including length constraints and word equations, cannot be entirely handled by automata-based solvers.

### 2019-11-22 evaluation.tex
related work and minor update to evalutaion

We compare {\tool} with other state-of-the-art string solvers, namely CVC4~(relese-1.7)~\cite{cvc4Tool}, Z3~(\texttt{dde8da853})~\cite{z3}, \textsf{Z3-str3}~(\texttt{109dfc5e8})~\cite{zheng2013z3}, \textsf{Trau+}~\cite{abdulla2019chain}, , and Ostrich~\cite{chen2019decision}. We do not compare with Sloth \cite{sloth} since it does not support length constraints, which occurs in most of our benchmarks. We also do not compare with ABC~\cite{aydin2018parameterized} (a model counter for string constraints), which over-approximates cases that cannot be modeled by \emph{aligned multi-track automata}~\cite{yu2016optimal} and hence cannot solve our benchmark examples precisely.

>>>>

We compare {\tool} with other state-of-the-art string solvers, namely CVC4~(relese-1.7)~\cite{cvc4Tool}, Z3~(\texttt{dde8da853})~\cite{z3}, \textsf{Z3-str3}~(\texttt{109dfc5e8})~\cite{zheng2013z3}, and \textsf{Trau+}~\cite{abdulla2019chain}. We do not compare with Sloth \cite{sloth} since it does not support length constraints, which occurs in most of our benchmarks. We also do not compare with ABC~\cite{aydin2018parameterized} (a model counter for string constraints) and Ostrich~\cite{chen2019decision}, because most of our benchmarks contain non-trivial equality constraints, which is not fully supported by those solvers.
